Qualitative Probabilistic Programming

### 2015-06-05 paper/paper.tex
cleanup

  It is written as a generative model for the observations.  Notice that
  we declared two types (\texttt{Point} and \texttt{Cluster}) and one
  random function \texttt{getPoint}).  Type annotations are necessary for random
  functions.  The type \texttt{Vector(2, Double)} expresses the fact that the
  points are 2-dimensional, and the type \texttt{Categorical(2)} expresses the
  fact that there are 2 possible clusters (so a cluster may be either 0 or 1).
  We use the \texttt{randomValue} function to generate a random uniform cluster. We do not know
  the distribution of points in each cluster, represented by the fact that \texttt{getPoint}
  is an unknown random function.  We will fill in the
  \texttt{getPoint} function with a learned function that will take a random
  sample from the given cluster.  The \texttt{observe} function allows us to
  observe data; here it says that our observations consist of \texttt{getPoint(cluster)} for
  each cluster.

@@

  Note that, because the two clusters are forced to have the same variance, they do not fit the data well,
  since the data has a different shape in each location.  To fix this problem, we can substitute the following
  model, which uses a separate random function for each cluster:
  \begin{lstlisting}
var Cluster = Categorical(2);
var Point = Vector(2, Double);
var getPointFunctions = [randFunction(Point), randFunction(Point)];

  Note that the clusters represent axis-aligned multivariate Gaussian distributions.   To allow them to be arbitrary Gaussian distributions (where the $x$ and $y$ coordinates may be correlated),
  we can use the following model:

  Here, there are 2 random functions for each cluster, one to get the $x$ coordinate and one to get the $y$ coordinate (whose distribution may depend on the $x$ coordinate).  This allows $x$ and $y$ to be correlated, improving fit:

@@

    \texttt{randFunction} returns a randomized function that is a member of some generalized linear model
    determined by the desired argument types and return type.
    The distribution
    of the function's return value $y$
    is some exponential family whose natural
    parameters are determined from the arguments $x$:

@@

    To determine $\eta(x)$, we label
    some subset of the sufficient statistics of both $x$ and $y$ as \emph{features}.  For the gaussian
    distribution, the sufficient statistics are $X$ and $X^2$ but the only feature is $X$.  For the
    categorical distribution \texttt{Categorical(n)}, the sufficient statistics
    and features are both $[X = 1], [X = 2] ..., [X=n - 1]$.
    The natural
    parameters corresponding to non-features are constant, while natural
    parameters corresponding to features are determined as an affine
    function of the features of the arguments.

@@

    where $\mathbf{N}$ is a matrix containing our parameters.  It must have 0 for each entry whose row corresponds
    to a sufficient statistics of $y$ that is not a feature and whose column is not 1.
    This ensures that only the
    natural parameters that are features of $y$ are affected by $x$.

@@

      \hline
    \texttt{Tuple(T1, T2)} & $\begin{bmatrix} \phi_{T1}(X_1) \\ \phi_{T2}(X_2) \end{bmatrix}$
                           & $\begin{bmatrix} \psi_{T1}(X_1) \\ \psi_{T2}(X_2) \end{bmatrix}$ &
      Independent prediction of each component
      \\
      \hline
    \end{tabular}

@@

    For the expectation step, we must estimate latent variable distributions given
    fixed values for the parameters.  To do this, we can run the Quipp program
    with random functions set to use these fixed parameter values to generate their results.
    We use the Metropolis Hastings algorithm to perform inference in this program,
    yielding traces.  Next, for each random function, we can find all calls
    to it in the trace to get the training data.

@@

    Given $(x, y)$ samples, parameter estimation to maximize log probability is a convex
    problem because the log probability function is concave:
    $$\log p_{\mathbf{N}}(y | x) = \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix} ^T \mathbf{N} \phi(y) - g\left(\mathbf{N}^T \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix}\right)$$

>>>>

  We declared two types (\texttt{Point} and \texttt{Cluster}) and one
  random function (\texttt{getPoint}).  Type annotations are necessary for random
  functions.  The type \texttt{Vector(2, Double)} expresses the fact that the
  points are 2-dimensional, and the type \texttt{Categorical(2)} expresses the
  fact that there are 2 possible clusters (so a cluster may be either 0 or 1).

  The variable \texttt{model} specifies a generative model for a single data point.
  We use the \texttt{randomValue} function to generate a random uniform cluster, and then
  use \texttt{getPoint} to generate the point given the cluster.  Since \texttt{getPoint}
  is an unknown random function, we will need to infer its parameters.
  The \texttt{observe} function allows us to observe data; here it says that our
  observation consist of the result of the call \texttt{getPoint(cluster)}.

@@

@@

  In the previous example, we that \texttt{randFunction(Categorical(2), Vector(2, Double))} represented 2 axis-aligned Gaussian clusters with equal variances.  In general, \texttt{randFunction} returns a randomized function that is a member of some generalized linear model
    determined by the desired argument types and return type.
    The distribution
    of the function's return value $y$
    is some exponential family whose natural
    parameters are determined from the arguments $x$:

@@

    To determine $\eta(x)$, we label
    some subset of the sufficient statistics of both $x$ and $y$ as \emph{features}.  For the gaussian
    distribution, the sufficient statistics are $X$ and $X^2$ but the only feature is $X$.  For the
    categorical distribution \texttt{Categorical(n)}, the sufficient statistics
    and features are both $[X = 1], [X = 2], ..., [X=n - 1]$.
    The natural
    parameters corresponding to non-features are constant, while natural
    parameters corresponding to features are determined as an affine
    function of the features of the arguments.

@@

    where $\mathbf{N}$ is a matrix containing our parameters.  It must have 0 for each entry whose row corresponds
    to a sufficient statistics of $y$ that is not a feature and whose column is not 1.
    This ensures that only the
    natural parameters that are features of $y$ are affected by $x$.  The following shows the types
    in Quipp and their corresponding sufficient statistics, features, and English descriptions.

@@

      \hline
    \texttt{Tuple(T1, T2)} & $\begin{bmatrix} \phi_{T1}(X_1) \\ \phi_{T2}(X_2) \end{bmatrix}$
                           & $\begin{bmatrix} \psi_{T1}(X_1) \\ \psi_{T2}(X_2) \end{bmatrix}$ &
      Independent regression of each component
      \\
      \hline
    \end{tabular}

    Note that $\texttt{Vector(n, T)}$ is shorthand for $Tuple(T, T, ..., T)$, with $n$ copies of $T$.

  \section{Example: improving the clustering model}
  Using this knowledge about the form of the random functions, we can improve the clustering model from before.
  As observed in the previous graph, the two clusters are forced to have the
  same variance. They do not fit the data well, since the data has a different
  shape in each location.  To fix this problem, we can substitute the following
  model, which uses a separate random function (and therefore a separate axis-aligned Gaussian distribution) for each cluster:
  \begin{lstlisting}
var Cluster = Categorical(2);
var Point = Vector(2, Double);
var getPointFunctions = [randFunction(Point), randFunction(Point)];

  The fact that the gaussian distribution for each cluster must be axis-aligned limits the degree to which the model can fit the data.   To allow each cluster to be an arbitrary Gaussian distributions (where the $x$ and $y$ coordinates may be correlated),
  we can use the following model:

  Here, there are 2 random functions for each cluster, one to get the $x$ coordinate and one to get the $y$ coordinate (whose distribution may depend linearly on the $x$ coordinate).  This allows $x$ and $y$ to be correlated, improving fit:

@@

    For the expectation step, we must estimate latent variable distributions given
    fixed values for the parameters.  To do this, we can replace unknown random functions in the Quipp program
    with random functions set to use these fixed parameter values, yielding a probabilistic program.
    We use the Metropolis Hastings algorithm to perform inference in this program,
    yielding a distribution of execution traces (where each execution trace specifies the result of every
    call to a random function).  Next, for each random function, we can find all calls
    to it in the trace to get the training data.

@@

    Given $(x, y)$ samples, parameter estimation to maximize log probability is a convex
    problem because the log probability function is concave as a function of $\mathbf{N}$:
    $$\log p_{\mathbf{N}}(y | x) = \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix} ^T \mathbf{N} \phi(y) - g\left(\mathbf{N}^T \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix}\right)$$
